---
title: "EM Report"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
library(MASS)        # for glm.nb
library(tidyverse)   # general data tidying and wrangling
library(readxl)      # read from excel
library(stringr)     # string functions
library(pscl)        # Zero-inflation negative binomial regression
library(magrittr)    # loads "%>%"
library(knitr)

##### Data #####
emData <- read_excel('EM Data (2016-2017).xlsx') %>%
          mutate(Value = round(Value), # zeroinfl requires integers for Value - might be on a different scale
                 Loc = {sapply(`Sampling Location`, str_split, ' -> ') %>%
                        sapply(`[`, 2)}) %>% # several models have issues with number of locations - do by room
          filter(!is.na(Value))

# remove spaces from names
names(emData) <- str_replace_all(names(emData), ' ', '')

# test methods and site classes
testMethods <- unique(emData$TestMethod)
siteClass <- c('100', '10,000', '100,000') # unique(emData$SiteClassDesc)
```


```{r analysis, include=FALSE}
#' function to calcualte the critical levels
#'
#' @details
#' This function assumes emData is in the working environment that critLevels() is called from.
#'
#' @param m Test method
#' @param c siteClass
#' @param alpha1 Alpha level for Alert
#' @param alpha2 Alpha level for Action
#' @param vuongTest Logical flag to enable running the Vuong test to verify that the zero-invlated model is better than standard negative binomial regression
#' 
#' @return Returns critical levels for site class c for the test, m at two levels: alpha1 and alpha2.
critLevels <- function(m, c, nsamp = 15000, alpha1 = 0.95, alpha2 = 0.99, vuongTest = FALSE)
{
    dat <- filter(emData, TestMethod == m & SiteClassDesc == c)

    condLoc <- TRUE
    model0 <- try({zeroinfl(Value ~ 1 | Loc, dist = 'negbin', data = dat) %>%
                   summary() %>%
                   coefficients() %>%
                   `[[`('count')}, silent = TRUE)
        
    # if the this model fails, Loc is probably the problem (and thus probably not important for the model)
    if(class(model0) == 'try-error')
    {
        condLoc <- FALSE
        model0 <- zeroinfl(Value ~ 1, dist = 'negbin', data = dat) %>%
                  summary() %>%
                  coefficients() %>%
                  `[[`('count')
    }
        
    # check to see if zeroinfl is a better model than glm.nb
    # seems to be the case for our data...
    # this can take awhile, so we won't run it by default
    if(vuongTest)
    {
        null0 <- glm.nb(Value ~ 1, data = dat)
    
        vuong(model0, null0) # low p-value indicates Zero Inflation model is better
    }

    return(c(crit95 = exp(model0[1,1] + qnorm(alpha1)*model0[1,2]),
             crit99 = exp(model0[1,1] + qnorm(alpha2)*model0[1,2]),
             mean = exp(model0[1,1])))
}

if(file.exists('mainTable.RData'))
{
    load('mainTable.RData')
}else{
    # this contains our main table of results
    mainTable <- data_frame(method = rep(testMethods, each = 3),
                            class = rep(siteClass, 3),
                            currAlert = c(1760, 176000, 1760000, NA, 5, 50, 0, 5, 25),
                            currAction = c(3520, 352000, 3520000, 1, 10, 100, 1, 10, 50),
                            crit95 = NA,
                            crit99 = NA,
                            N = unlist(map2(method, class, ~ {filter(emData, TestMethod == .x &
                                                                     SiteClassDesc == .y)$Value %>%
                                                              length()})),
                            mean = NA)

    # temporary for loop to get long part of analysis finished
    # --- modify this line to run specific method/class results
    for(i in 1:dim(mainTable)[1])
    {
        mainTable[i,c(5:6, 8)] <- critLevels(mainTable$method[i], mainTable$class[i])
    }

    save(mainTable, file = 'mainTable.RData')
}
```

```{r mainTable, echo=FALSE}
select(mainTable, -N) %>%
    kable(digits = c(rep(0, 4), 3, 3, 3), align = c('l', rep('r', 6)))
```

The table above gives analysis results of historical environmental monitoring data and contains the following columns:

 * method - one of three tests: non-viable air, surface contact and viable air.
 * class - the level of expected cleanliness of the room, with lower numbers expected to have fewer environmental contaminants.
 * currAlert - the current threshold which, if exceeded, will trigger an alert.
 * currAction - the current threshold which, if exceeded, will trigger an action to remediate the exception.
 * crit95 - the threshold at which 95% of historical observations have fallen below as calculated by a zero-inflated negative binomial model^†^.
 * crit99 - the threshold at which 99% of historical observations have fallen below as calculated by a zero-inflated negative binomial model^†^.
 * mean - the mean level of historical observations as calculated by a zero-inflated negative binomial model^†^.
 
† The zero-inflated negative binomial model accounts for over-dispersion of observations that results in applications where there are an excess of zeros.

\newline

\newline

```{r negbin model, echo=FALSE}

mainTable <- mutate(mainTable,
                    crit95_negbin = NA,
                    crit99_negbin = NA,
                    mean_negbin = NA)

for(i in 1:dim(mainTable)[1])
{
    null0 <- suppressWarnings(glm.nb(Value ~ 1, data = filter(emData, TestMethod == mainTable$method[i] & 
                                                              SiteClassDesc == mainTable$class[i]))) %>%
             summary() %>%
             `[[`('coefficients')
    
    mainTable$crit95_negbin[i] <- exp(null0[1,1] + null0[1,2]*qnorm(0.95))
    mainTable$crit99_negbin[i] <- exp(null0[1,1] + null0[1,2]*qnorm(0.99))
    mainTable$mean_negbin[i] <- exp(null0[1,1])
}

select(mainTable, -N, -crit95, -crit99, -mean) %>%
    rename(crit95 = crit95_negbin, crit99 = crit99_negbin, mean = mean_negbin) %>%
    kable(digits = c(rep(0, 4), 3, 3, 3), aligh = c('l', rep('r', 6)))
```

This second table gives analysis results for a standard negative binomial model of historical environmental monitoring data and contains the following columns:

  * method - one of three tests: non-viable air, surface contact and viable air.
  * class - the level of expected cleanliness of the room, with lower numbers expected to have fewer environmental contaminants.
  * currAlert - the current threshold which, if exceeded, will trigger an alert.
  * currAction - the current threshold which, if exceeded, will trigger an action to remediate the exception.
  * crit95 - the threshold at which 95% of historical observations have fallen below as estimated by a negative binomial model.
  * crit99 - the threshold at which 99% of historical observations have fallen below as estimated by a negative binomial model.
  * mean - the mean level of historical observations as estimated by a negative binomial model.
  