---
title: "EM Report"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
library(MASS)        # for glm.nb
library(tidyverse)   # general data tidying and wrangling
library(readxl)      # read from excel
library(stringr)     # string functions
library(pscl)        # Zero-inflation negative binomial regression
library(parallel)    # allows running boot functions in parallel
library(boot)        # does the boot strap simulations
library(magrittr)    # loads "%>%"
library(knitr)

##### Data #####
emData <- read_excel('EM Data (2016-2017).xlsx') %>%
          mutate(Value = round(Value), # zeroinfl requires integers for Value - might be on a different scale
                 Loc = {sapply(`Sampling Location`, str_split, ' -> ') %>%
                        sapply(`[`, 2)}) %>% # several models have issues with number of locations - do by room
          filter(!is.na(Value))

# remove spaces from names
names(emData) <- str_replace_all(names(emData), ' ', '')

# test methods and site classes
testMethods <- unique(emData$TestMethod)
siteClass <- c('100', '10,000', '100,000') # unique(emData$SiteClassDesc)
```


```{r analysis, include=FALSE}
#' function to calcualte the critical levels
#'
#' @details
#' This function assumes emData is in the working environment that critLevels() is called from.
#'
#' @param m Test method
#' @param c siteClass
#' @param alpha1 Alpha level for Alert
#' @param alpha2 Alpha level for Action
#' @param vuongTest Logical flag to enable running the Vuong test to verify that the zero-invlated model is better than standard negative binomial regression
#' 
#' @return Returns critical levels for site class c for the test, m at two levels: alpha1 and alpha2.
critLevels <- function(m, c, nsamp = 15000, alpha1 = 0.95, alpha2 = 0.99, vuongTest = FALSE)
{
    if(file.exists(paste0('boot_', m, '_', c, '.RData')))
    {
        load(paste0('boot_', m, '_', c, '.RData'))
    }else{
        dat <- filter(emData, TestMethod == m & SiteClassDesc == c)

        condLoc <- TRUE
        model0 <- try(zeroinfl(Value ~ 1 | Loc, dist = 'negbin', data = dat), silent = TRUE)
        
        # if the this model fails, Loc is probably the problem (and thus probably not important for the model)
        if(class(model0) == 'try-error')
        {
            condLoc <- FALSE
            model0 <- zeroinfl(Value ~ 1, dist = 'negbin', data = dat)
        }
        
        # check to see if zeroinfl is a better model than glm.nb
        # seems to be the case for our data...
        # this can take awhile, so we won't run it by default
        if(vuongTest)
        {
            null0 <- glm.nb(Value ~ 1, data = dat)
    
            vuong(model0, null0) # low p-value indicates Zero Inflation model is better
        }
    
        # this function will run one bootstrap sample & return the count statistic
        boot_ci <- function(dat, i, counts, zeros, condLoc)
        {
            require(pscl)
            require(magrittr)
            if(condLoc)
            {
                retval <- try({
                    zeroinfl(Value ~ 1 | Loc, dist = 'negbin',
                            data = dat[i,], # start with one randomly picked row
                            start = list(count = counts,
                                         zero = zeros)) %>%
                        summary() %>%        # summary object
                        coef() %>%           # get coefficients
                        `[[`(1) %>%          # first item is count estimates
                        `[`(1,1)             # first row/first column = estimate
                }, silent = TRUE)
            }else{
                retval <- try({
                    zeroinfl(Value ~ 1, dist = 'negbin',
                            data = dat[i,], # start with one randomly picked row
                            start = list(count = counts,
                                         zero = zeros)) %>%
                        summary() %>%        # summary object
                        coef() %>%           # get coefficients
                        `[[`(1) %>%          # first item is count estimates
                        `[`(1,1)             # first row/first column = estimate
                }, silent = TRUE)
            }
        
            if(class(retval) == 'try-error')
                return(NA)
        
            return(retval)
        }
    
        # bootstrap samples
        set.seed(389427)
        bs_samples <- boot(dat, boot_ci, stype = 'i', R = nsamp, 
                           parallel = 'snow', ncpus = detectCores(),
                           counts = coef(model0, 'count'),
                           zeros = coef(model0, 'zero'),
                           condLoc = condLoc)
        
        # some of these return NA for some reason - drop these
        todrop <- which(is.na(bs_samples$t))
        
        if(length(todrop) > 0)
        {
            bs_samples$t <- bs_samples$t[-todrop,,drop = FALSE]
            bs_samples$R <- length(bs_samples$t)
        }
        
        # save this for future runs!
        save(bs_samples, file = paste0('boot_', m, '_', c, '.RData'))
    }

    # calculate critical levels
    btci <- suppressWarnings(boot.ci(bs_samples, conf = c(alpha1, alpha2)))
    
    return(c(crit95 = exp(btci$basic[1,5]),
             crit99 = exp(btci$basic[2,5]),
             mean = exp(btci$t0)))
}

if(file.exists('mainTable.RData'))
{
    load('mainTable.RData')
}else{
    # this contains our main table of results
    mainTable <- data_frame(method = rep(testMethods, each = 3),
                            class = rep(siteClass, 3),
                            currAlert = c(1760, 176000, 1760000, NA, 5, 50, 0, 5, 25),
                            currAction = c(3520, 352000, 3520000, 1, 10, 100, 1, 10, 50),
                            crit95 = NA,
                            crit99 = NA,
                            N = unlist(map2(method, class, ~ {filter(emData, TestMethod == .x &
                                                                     SiteClassDesc == .y)$Value %>%
                                                              length()})),
                            mean = NA)

    # temporary for loop to get long part of analysis finished
    # --- modify this line to run specific method/class results
    for(i in 1:dim(mainTable)[1])
    {
        mainTable[i,c(5:6, 8)] <- critLevels(mainTable$method[i], mainTable$class[i])
    }

    save(mainTable, file = 'mainTable.RData')
}
```

```{r mainTable, echo=FALSE}
# mutate(mainTable,
#        crit95 = ifelse(crit95 <   1, round(crit95, 3), 
#                 ifelse(crit95 < 100, round(crit95, 1),
#                                      round(crit95))),
#        crit99 = ifelse(crit99 <   1, round(crit99, 3),
#                 ifelse(crit99 < 100, round(crit99, 1),
#                                      round(crit99))),
#        mean   = ifelse(mean   <   1, round(mean, 3),
#                 ifelse(mean   < 100, round(mean, 1),
#                                      round(mean)))) %>%
select(mainTable, -N) %>%
    kable(digits = c(rep(0, 4), 3, 3, 0, 3), align = c('l', rep('r', 7)))
```

The table above gives analysis results of historical environmental monitoring data and contains the following columns:

 * method - one of three tests: non-viable air, surface contact and viable air.
 * class - the level of expected cleanliness of the room, with lower numbers expected to have fewer environmental contaminants.
 * currAlert - the current threshold which, if exceeded, will trigger an alert.
 * currAction - the current threshold which, if exceeded, will trigger an action to remediate the exception.
 * crit95 - the threshold at which 95% of historical observations have fallen below as calculated by a zero-inflated negative binomial model^†^.
 * crit99 - the threshold at which 99% of historical observations have fallen below as calculated by a zero-inflated negative binomial model^†^.
 * mean - the mean level of historical observations as calculated by a zero-inflated negative binomial model^†^.
 
† The zero-inflated negative binomial model accounts for over-dispersion of observations that results in applications where there are an excess of zeros.